---
title: "Gene Annotations"
output: 
  github_document:
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.retina = 2)
library(tidyverse)
```

As a precursor to a range of gene functional analyses, we generated an annotated gene set for *A. digitifera*.  Our annotations are based on gene models provided by Dr. Chuya Shinzato for the version 2 genome. These annotations were provided as file `Adig-v2.gff` along with a translation table `20191203c.shinzato-0010_scaff_adig.acclist.txt` to convert scaffold ids in this file to those in the NCBI assembly, `GCA_014634065.1_Adig_2.0_genomic.fna`.


First we extracted protein coding sequences and nucleotide sequences for the longest isoform per gene model. These sequences were used for BLAST and InterProScan analyses. 

```bash
cgat gff2gff --filter=longest-gene -I annotation_from_chuya/adig-v2.gff -S annotation_longest_gene.gff

gffread -g reference_entryname.fa -y protein.fa annotation_longest_gene.gff
gffread -g reference_entryname.fa -x CDS.fa annotation_longest_gene.gff
```

## Scan for conserved domains

Next we used [InterProScan5](https://www.ebi.ac.uk/interpro/search/sequence/) version `5.53-87` to identify conserved domains in protein translations of all genes.  Prior to running this scan any non-standard amino acid characters (ambiguities or premature stop codons) in protein sequences were removed. This analysis was primarily used to provide a set of GO term assignments based on conserved domains rather than specific genes.  We call these GO terms `ipr_go` to distinguish them from those obtained from uniprot, which are terms assigned to a specific homologous gene. The `ipr_go` terms will tend to be less specific but are likely to be more reliable than those provided by uniprot. 

Interproscan was invoked in batches of 1000 sequences as follows. 

```bash
interproscan.sh -i $seqs --disable-precalc -goterms
```

The `tsv` files produced by this process were concatenated to produce a single file which we include here as `data/hpc/annotation/all_ipr.tsv`

```{r}
# We extract GO term assignments from Interproscan results.
ipr_colnames <- c("adi_id","md5","seqlen","analysis","signature_acc","signature_desc","start","stop","score","status","date","ipr_acc","ipr_ann","ipr_go", "pathways")

interpro2 <- read_tsv("data/hpc/annotation/all_ipr.tsv", guess_max = 10000, col_names = ipr_colnames, show_col_types = FALSE) %>% 
  dplyr::select(adi_id,ipr_go) %>% 
  filter(ipr_go!="-") %>% 
  na.omit()

interpro_go <- interpro2 %>% 
  separate_rows(ipr_go,sep="\\|") %>% 
  mutate(geneid=str_remove(adi_id,"\\.t\\d")) %>% 
  unique %>% 
  group_by(geneid) %>% 
  dplyr::summarise(go=paste(ipr_go,collapse = "; "))
```

## Homology search with BLAST

To identify homologs of *A. digitifera* with high quality functional annotations we used BLAST[xp] to search all genes against the swissprot uniprot database. After filtering blast results to include only those with evalue <1e-5, we then selected the best available hit based on evalue.  For all these best hits we then looked up putative gene names, GO terms, and Kegg information from [Uniprot ID mapping](https://www.uniprot.org/uploadlists/) from UniprotKB AC/ID to UniprotKB.

```{r generate_annot_table}
bl6_cols <- c("qaccver","saccver","pident","length","mismatch","gapopen","qstart","qend","sstart","send","evalue","bitscore")

blastp <- read_tsv("data/hpc/annotation/Adigitifera_uniprot_blastp.outfmt6", col_names = bl6_cols) %>% 
  dplyr::select(adi_id=qaccver, uniprot_id=saccver,evalue) %>% 
  add_column(method="blastp") %>% 
  mutate(geneid=str_remove(adi_id,'\\.t\\d'))

blastx <- read_tsv("data/hpc/annotation/Adigitifera_uniprot_blastx.outfmt6", col_names = bl6_cols) %>% 
  dplyr::select(adi_id=qaccver, uniprot_id=saccver,evalue) %>% 
  add_column(method="blastx") %>% 
  mutate(geneid=str_remove(adi_id,'\\.t\\d'))

# This file was obtained using the uniprot mapping service
uniprotkb_tab <- read_tsv("data/hpc/annotation/uniprot-yourlist.tab") %>% dplyr::select(-Status)

uniprot_gene_annot <- rbind(blastp,blastx) %>% 
  group_by(geneid) %>%
  slice_min(order_by = evalue, n = 1, with_ties = FALSE) %>% 
  left_join(uniprotkb_tab,by=c("uniprot_id"="Entry")) %>%
  ungroup %>% 
  dplyr::select(geneid,uniprot_id,entryname="Entry name",
          genename="Gene names",go="Gene ontology IDs",kegg="Cross-reference (KEGG)",protein="Protein names") %>% mutate(genename=ifelse(is.na(genename),entryname,genename)) %>% 
  left_join((interpro_go %>% dplyr::rename(go_ipr=go)),by=c("geneid"))




```


```{r}
# Write out results to a table.  This can easily be used in other analyses for GO enrichment etc.
write_tsv(uniprot_gene_annot, "data/hpc/annotation/uniprot_gene_annot.tsv")
```


```{r annotation-res-stats-table}
all <- read_tsv("data/hpc/annotation/allgenes_ids_extracted_from_gff.txt",col_names = "geneid")
all_annot<- all %>% left_join(interpro_go) %>% 
  left_join(uniprot_gene_annot %>% dplyr::select(geneid,go),by="geneid") %>% 
  dplyr::rename("InterProScan"=go.x,"Uniprot(blast)"=go.y) 

all_summ <- map_df(all_annot %>% dplyr::select(InterProScan,`Uniprot(blast)`), ~sum(!is.na(.))/nrow(all_annot)) 
```

## Summary

A final table of annotated genes is provided as the file `data/hpc/annotation/uniprot_gene_annot.tsv` available as part of the data package for this repository. Overall we found that `r all_summ$all_summ$InterProScan` percent of genes could be annotated with Interproscan while `r all_summ[,2]` could be annotated with a blast hit to Swissprot.
