---
title: "Gene Ontology Enrichment analysis"
output: 
  github_document:
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.retina = 2)
library(tidyverse)
library(RColorBrewer)
require(topGO)
require(ggsci)
library(aplot)
library(ape)
library(GO.db)
library(ggtree)
library(ggstance)
library(cowplot)
```


```{r}
# If not present run 09.annotate
uniprot_gene_annot <- read_tsv("data/hpc/annotation/uniprot_gene_annot.tsv", show_col_types = FALSE) %>% 
  mutate(go=go_ipr) %>% 
  dplyr::select(-go_ipr)
```


```{r}
read_sg <- function(path){
  read_tsv(path,col_names = NULL,show_col_types = FALSE) %>% 
  unite("region_id",X10,X13,X14,sep="_") %>% 
  tidyr::extract(X9,into="gene_id",regex="ID=(.*)") %>% 
  dplyr::select(region_id,gene_id)
}

sweep_genes <- rbind(
  read_sg("data/hpc/selection2/pbs/pbs_genes_in_sweeps.tsv") %>% add_column(pop="inshore"),
  read_sg("data/hpc/selection2/pbs/pbs_genes_no_sweeps.tsv") %>% add_column(pop="northoffshore"),
  read_sg("data/hpc/selection2/pbs/pbs_genes_so_sweeps.tsv") %>% add_column(pop="southoffshore")
)
```




```{r}
candidate_genes_table <- sweep_genes %>% 
  left_join(uniprot_gene_annot,by=c("gene_id"="geneid"))
```

In a previous analysis (see [14.ehh_pbs](14.ehh_pbs.md) we identified a total of  `r nrow(candidate_genes_table)` genes that overlapped with `r candidate_genes_table$region_id %>% n_distinct()` regions identified as being in the top 1% of EHH stats and also as having significant PBS values. A summary of sweep regions and the number of genes covered by them in each population is shown below;

```{r}
candidate_genes_table %>% 
  group_by(pop) %>% 
  dplyr::summarise(numgenes = n(), nregions = length(unique(region_id))) %>% 
  knitr::kable()
```

We used `topGO` to test whether any GO terms were enriched in the gene sets covered by selective sweeps in each population. Enrichment was initially calculated in a gene-centric manner because GO terms are most properly interpreted as attached to genes not regions.  This initial (naive) enrichment calculation identified several significantly enriched GO terms, however, as can be seen in the table below these were almost always supported by just a single region.  


```{r define-functions-regions}
# If not present run 10.identify
regions_genes_go <- read_rds("cache/all_regions_genes.rds") %>% 
  unite("region_id",chr,start,end) %>% 
  left_join(uniprot_gene_annot,by="geneid")

regions2go <- regions_genes_go %>% 
  group_by(region_id) %>%
  dplyr::summarise(go = paste(go,sep=";")) %>% 
  split(.$region_id) %>% 
  purrr::map(~str_split(.$go,";") %>% pluck(1) %>% str_trim() %>% unique())
```


```{r define-functions-genes}
gene2go <- uniprot_gene_annot %>% 
  split(.$geneid) %>% 
  purrr::map(~str_split(.$go,";") %>% pluck(1) %>% str_trim())

go2gene <- uniprot_gene_annot %>% 
  dplyr::select(geneid,go) %>% 
  separate_rows(go,sep=";") %>% 
  mutate(go=str_trim(go)) %>% 
  group_by(go) %>%
  split(.$go) %>% 
  purrr::map( ~.$geneid)

go_enrich_topgo <- function(targets,onto) {
    genenames <- names(gene2go)
    genelist <- factor(as.integer(genenames %in% targets))
    names(genelist) <- genenames

    GOdata <- new("topGOdata",
                  allGenes = genelist,
                  annot=annFUN.gene2GO,
                  gene2GO=gene2go,
                  ontology=onto,
                  nodeSize=10)
    
    resultFisher <- runTest(GOdata, 
                        algorithm = "weight01",
                        statistic = "fisher")

    gt <- GenTable(GOdata,classic=resultFisher,
                   WeightFisher=resultFisher,
                   orderBy = "WeightFisher",
                   topNodes=150)
    list(godata= GOdata, result=resultFisher, table=gt)
}

genes_in_enrichedGO <- function(topgo_results,target) {
  topgo_results$table %>% 
    mutate(classic=as.numeric(classic),
           WeightFisher=as.numeric(WeightFisher)) %>%
    dplyr::select(GO.ID) %>% pull %>% 
    map_df(~genesInTerm(object = topgo_results$godata,whichGO = .x)) %>%
    pivot_longer(cols = everything(),names_to = "go",values_to = "geneid") %>% 
    na.omit() %>% 
    filter(geneid %in% target)
}

go_enrich_all_ontologies <- function(genes_table,target_pop){
  target <- genes_table %>% 
    filter(pop==target_pop) %>% 
    dplyr::select(gene_id) %>% 
    distinct %>% 
    pull

  mf <- go_enrich_topgo(targets = target,onto = "MF")
  bp <- go_enrich_topgo(targets = target,onto = "BP")
  cc <- go_enrich_topgo(targets = target,onto = "CC")

  all_results <- list("MF"=mf,"BP"=bp,"CC"=cc)
  all_results_genes <- map_dfr(all_results,genes_in_enrichedGO,target) %>% 
    left_join(uniprot_gene_annot %>%
                dplyr::select(geneid,uniprot_id,protein)) %>%
    group_by(go) %>% 
    dplyr::summarise(uniprot_ids = paste(unique(uniprot_id),collapse = ";"), 
                geneids = paste(unique(geneid),collapse = ";"),
                protein_names = paste(unique(protein),collapse = ";"))
  
  all_onto <- rbind(bp$table %>% mutate(classic=as.numeric(classic),WeightFisher=as.numeric(WeightFisher)) %>% add_column(ontology="BP"), 
                    mf$table %>% mutate(classic=as.numeric(classic),WeightFisher=as.numeric(WeightFisher)) %>% add_column(ontology="MF"),
                    cc$table %>% mutate(classic=as.numeric(classic),WeightFisher=as.numeric(WeightFisher)) %>% add_column(ontology="CC"))
  all_onto %>% 
    left_join(all_results_genes,by=c("GO.ID"="go")) 
}
```


```{r}
if ( !file.exists("cache/topgo_results_all.rds") ){
  inshore_topgo <- go_enrich_all_ontologies(candidate_genes_table,"inshore") %>% add_column(pop="inshore")
  northoffshore_topgo <- go_enrich_all_ontologies(candidate_genes_table,"northoffshore") %>% add_column(pop="northoffshore")
  southoffshore_topgo <- go_enrich_all_ontologies(candidate_genes_table,"southoffshore") %>% add_column(pop="southoffshore")

  topgo_results_all <- rbind(inshore_topgo,northoffshore_topgo,southoffshore_topgo)
  write_rds(topgo_results_all,"cache/topgo_results_all.rds")
  write_tsv(topgo_results_all,"cache/topgo_results_all.tsv") # This table should be in Supp Info

  } else {
  topgo_results_all <- read_rds("cache/topgo_results_all.rds")
}
```


```{r}
jaccard <- function(a, b) {
    1-(length(intersect(a, b))/length(union(a,b)))
}


jaccard_go <- function(terms1){
  jaccard(terms1[[1]],terms1[[2]])
}

display_go <- topgo_results_all %>% 
  filter(Significant>0) %>% 
  pull(GO.ID) %>% 
  unique()

go2gene_signif <- topgo_results_all %>%
  filter(GO.ID %in% display_go) %>%
  dplyr::select(GO.ID,geneids) %>%
  separate_rows(geneids,sep=";") %>% 
  filter(!is.na(geneids)) %>% 
  filter(geneids!="NA") %>% 
  group_by(GO.ID) %>% 
  dplyr::summarise(geneids=paste(unique(geneids),collapse = ";")) %>% 
  ungroup() %>% 
  split(.$GO.ID) %>%
  purrr::map(~ .$geneids %>% str_split(";") %>% pluck(1))



if ( !file.exists("cache/jdm.rds")){
  cx <- cross2(go2gene_signif,go2gene_signif) #
  jdm <- cx %>% 
    map_dbl(jaccard_go) %>% 
    matrix(nrow=length(go2gene_signif))
  write_rds(jdm,"cache/jdm.rds")
} else {
  jdm <- read_rds("cache/jdm.rds")
}

rownames(jdm) <- names(go2gene_signif)
colnames(jdm) <- names(go2gene_signif)
```

```{r, fig.height=12}
n_regions_per_term <- function(term,focal_pop){
  candidate_genes_table %>%
  filter(gene_id %in% (go2gene_signif[term] %>% pluck(1))) %>% 
    filter(pop==focal_pop) %>% 
    pull(region_id) %>% n_distinct()
}

n_genes_per_term <- function(term,focal_pop){
  candidate_genes_table %>%
    filter(pop==focal_pop) %>% 
    filter(gene_id %in% (go2gene_signif[term] %>% pluck(1))) %>% 
    pull(gene_id) %>% n_distinct()
}

n_regions_bg_per_term <- function(term){
  regions_genes_go %>% 
    dplyr::select(region_id,geneid) %>% 
    filter(geneid %in% (go2gene_signif[term] %>% pluck(1))) %>% 
    n_distinct()
}

enrich_test <- function(n_regions,n_regions_sig,n_regions_ann,n_regions_ann_sig){
  dat <- data.frame(
  "non_signif" = c(n_regions_ann-n_regions_ann_sig, n_regions-n_regions_ann),
  "signif" = c(n_regions_ann_sig, n_regions_sig-n_regions_ann_sig),
  row.names = c("Annotated", "Non-Annotated"),
  stringsAsFactors = FALSE
  )
  colnames(dat) <- c("Non_Sig", "Sig")
  test <- fisher.test(dat)
  test$p.value
}


number_of_sig_regions <- candidate_genes_table %>% 
  dplyr::select(region_id,pop) %>% 
  group_by(pop) %>% 
  dplyr::summarise(n_sig = n()) %>% 
  pull(n_sig,name = pop)

total_bg_regions <- regions_genes_go$region_id %>% n_distinct()
```

```{r supp-table}
topgo_results_all %>% 
  filter(classic <0.005) %>% 
  mutate(n_regions = map2_int(GO.ID,pop,n_regions_per_term),
         n_region_bg = map_int(GO.ID,n_regions_bg_per_term),
         n_region_expected = number_of_sig_regions[pop]*n_region_bg/total_bg_regions) %>% 
  mutate(p.region = pmap_dbl(list(total_bg_regions,number_of_sig_regions[pop],n_region_bg,n_regions),enrich_test)) %>% 
  dplyr::select(-uniprot_ids,-protein_names,-n_region_expected) %>% 
  dplyr::rename("Num Sweeps"=n_regions,"Num Background Regions"=n_region_bg,"Region p-value"=p.region) %>% 
    knitr::kable()
```



